Supporting ecological decision making using feature-selection and variable importance

### 2019-06-30 code/98-paper/journal/manuscript.tex
optimize introduction and data

	\title{Hyperspectral tree defoliation modeling with a focus on feature-selection and model interpretation}
	%% Group authors per affiliation:
	\author[FSU]{Patrick Schratz}
	\cortext[mycorrespondingauthor]{Corresponding author}
	\ead{patrick.schratz@uni-jena.de}

@@

% Explain how remote sensing is used in forestry (potential to map forest health)
% Link remote sensing and machine-learning
\noindent The use of \ac{ML} algorithms for analyzing remote sensing data has seen a huge increase in the last decade \citep{lary2016}.
Especially since the launch of the first Sentinel satellite in 2014 the amount of freely available imagery data has increased greatly.
At the same time, the implementation and usability of algorithms (both statistical and machine-learning) has been greatly simplified in the open-source community.
Scientists can nowadays relatively easily feed large amounts of information into complex models in a semi-automated way.
These points enable the possibility to analyze (temporal) changes of the environment on a large scale \citep{ma2015}.

% link to forest health analysis and show exemplary studies
Machine-learning methods in combination with remote-sensing data are used in many environmental fields such as vegetation cover analysis or forest carbon storage mapping \citep{mascaro2014, urban2018}.
The advantages of conducting analysis at large scale make these points a promising toolset also for forest health analysis.
This research field aims to enhance the understanding of biotic and abiotic triggers, for example by analyzing defoliation at trees \citep{hawrylo2018}.
Different approaches exist including temporal change detections \citep{zhang2016} or analysis about the current health status of forests on a stand level \citep{townsend2012}.
Defoliation at trees is an information which is able to serve as a proxy for forest health by describing the impact of biotic and abiotic pest triggers \citep{townsend2012, goodbody2018}.
Vegetation indices have shown the potential to provide valuable information when analyzing such scenarios \citep{jiang2014, adamczyk2015}.
Since both triggers and tree types are highly variable across regions, it cannot be known apriori which indices are most important for describing the actual relationship.
This emphasizes the need to extract as much information as possible from the available input data to generate features that can possibly explain the modeled relationship.
Besides \ac{VI} a less known type of indices that can be calculated from remote-sensing products are \ac{NRI}.
\ac{NRI} are arbitrary combinations of band information from remote-sensing data.
Especially when working with hyperspectral data, \ac{NRI} provide the possibility to calculate hundreds of indices that can be used during modeling.

@@

Despite its popularity in environmental modeling, there are no studies so far that used machine-learning algorithms in combination with remote-sensing data to analyze defoliation on a tree level.
This study aims at closing this gap by analyzing defoliation at trees in northern Spain using airborne hyperspectral data.
To cope with the state-of-the-art methodology in modeling, a methodology which combines feature-selection and hyperparameter tuning across multiple algorithms is used.
The approach of generating many features comes along with the introduction of "high-dimensionality" into the analysis \citep{trunk1979, xu2016}.
Even though \ac{ML} algorithms are capable of handling highly-correlated input variables, the fitting time of models increase substantially and the interpretation of results becomes more complicated.
Besides modeling defoliation at trees this work also aims at providing an exemplary analysis of how to deal with spatial feature-rich data derived from a hyperspectral remote-sensing source.

@@

\begin{itemize}
	\item Do different environmental feature sets show substantial differences in performance when predicting defoliation at trees?
	\item How are different feature-selection methods performing and do they improve the predictive performance of the final model?
	\item Is it possible to map defoliation at a tree level using high-resolution hyperspectral data at a sufficient error rate?
	\item Which features were most important to the model to detect defoliation at trees?\end{itemize}

\noindent State-of-the-art machine-learning techniques were compared using three feature sets with different feature-selection approaches including a relatively novel approach of ensemble filters.
Model-based optimization was applied for hyperparameter tuning of the chosen algorithms.
Spatial block \ac{CV} was used to account for spatial autocorrelation in the data.

\section{Data and study area}
\noindent Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain.
The trees in the study area plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum} leading to a spread of cankers or defoliation \citep{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} spanning a range from 0\% - 100\% (\autoref{fig:defol-distr}).
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \citep{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \citep{ganley2009}.

\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width=0.7\textwidth] {defoliation-distribution-plot-1.pdf}}
		\caption{Distribution of defoliation at trees for plots Laukiz 1, Laukiz 2, Luiando and Oiartzun.}
		\label{fig:defol-distr}
	\end{center}
\end{figure}

@@

\subsection{Derivation of indices}
% TODO: Mention different feature sets
\noindent To use the full information from the hyperspectral data, all possible vegetation indices supported by the R package \textit{hsdar} (90 in total) as well as all possible \ac{NRI} combinations were calculated.
The following formula was used for the NRI calculation:

@@

% spectral signatures
%\begin{figure} [H]
%\begin{center}
%\makebox[\textwidth]{\includegraphics[width=\textwidth] {spectral_signatures.pdf}}
%\caption{Spectral signatures (mean and standard deviation) of each plot.}
%\label{fig:spectral_signatures}
%\end{center}
%\end{figure}

@@

>>>>

	\title{Supporting ecological decision making using feature-selection and variable importance}
	%% Group authors per affiliation:
	\author[FSU]{Patrick Schratz}
	\cortext[mycorrespondingauthor]{Corresponding author}
	\ead{patrick.schratz@uni-jena.de}

@@

% Explain how remote sensing is used in forestry (potential to map forest health)
% Link remote sensing and machine-learning
\noindent The use of \ac{ML} algorithms for analyzing remote sensing data has seen a huge increase in the last decade \citep{lary2016}.
Especially since the launch of the first Sentinel satellite in 2014 the amount of freely available imagery data has increased greatly.
At the same time, the implementation and usability of algorithms (both statistical and machine-learning) has been greatly simplified in the open-source community.
Scientists can nowadays relatively easily feed large amounts of (environmental) information into complex models and analyze the data in a semi-automated way.
This simplifies the analysis of (temporal) changes of the environment on a large scale \citep{ma2015}.

% link to forest health analysis and show exemplary studies
Machine-learning methods in combination with remote-sensing data are used in many environmental fields such as vegetation cover analysis or forest carbon storage mapping \citep{mascaro2014, urban2018}.
The ability of predicting to large unknown areas qualifies these tools as a promising toolset for forest health analysis.
This research field aims to enhance the understanding of biotic and abiotic triggers, for example by analyzing defoliation at trees \citep{hawrylo2018}.

Approaches for analyzing forest health include temporal change detections \citep{zhang2016} or describing the current health status of forests on a stand level \citep{townsend2012}.
In such studies, defoliation at trees is a variable able to serve as a proxy for forest health by describing the impact of biotic and abiotic pest triggers \citep{townsend2012, goodbody2018}.

Vegetation indices have shown the potential to provide valuable information when analyzing such scenarios \citep{jiang2014, adamczyk2015}.
Since both triggers and tree types are highly variable across regions, it cannot be known apriori which indices are most important for describing the actual relationship.
This emphasizes the need to extract as much information as possible from the available input data to generate features that can possibly explain the modeled relationship.
Besides \ac{VI} a less known index type which can be calculated from remote-sensing products is the \ac{NRI}.
\ac{NRI}s are arbitrary combinations of band information from remote-sensing data.
Especially when working with hyperspectral data, hundreds of \ac{NRI}s can be derived and used during modeling.

@@

Despite its popularity in environmental modeling, there are no studies so far that used machine-learning algorithms in combination with remote-sensing data to analyze defoliation on a tree level.
This study aims at closing this gap by analyzing defoliation at trees in northern Spain using airborne hyperspectral data.
To cope with the state-of-the-art methodology in modeling, a methodology which combines feature-selection and hyperparameter tuning across multiple algorithms was used.
The approach of generating many features comes along with the possible introduction of "high-dimensionality" into the analysis \citep{trunk1979, xu2016}.

Even though \ac{ML} algorithms are capable of handling highly-correlated input variables, the fitting time of models increase substantially and the interpretation of results becomes more complicated.
Besides modeling defoliation at trees and interpreting the model results from an ecological point of view, this work also aims at providing an exemplary analysis of how to deal with spatial feature-rich datasets.

@@

\begin{itemize}
	\item Do different environmental feature sets show substantial differences in performance when predicting defoliation at trees?
    \item How are various feature-selection methods influencing the predictive performance of the models?
	\item Which features are most important and how can these be interpreted in an ecological context?
\end{itemize}

\noindent State-of-the-art machine-learning techniques were compared using three feature sets across different feature-selection approaches including a relatively novel approach of "ensemble filters".
Model-based optimization was applied for hyperparameter tuning of the chosen algorithms which included an optimization of the feature set.
Spatial block \ac{CV} was used to account for spatial autocorrelation in the data.

\section{Data and study area}
% TODO: Add picture of affected trees

\noindent Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain.
The trees in the study area plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum} leading to a spread of cankers or defoliation \citep{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (serving as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} spanning a range from 0\% - 100\% (\autoref{fig:defol-distr}).
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \citep{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \citep{ganley2009}.

\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width=0.7\textwidth] {defoliation-distribution-plot-1.pdf}}
        \caption{Response variable defoliation at trees visualized as violin plots showing the variance of in-situ measurements for plots \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}.}
		\label{fig:defol-distr}
	\end{center}
\end{figure}

@@

\subsection{Derivation of indices}
\noindent To use the full information from the hyperspectral data, all possible vegetation indices supported by the R package \textit{hsdar} (89 in total) as well as all possible \ac{NRI} combinations were calculated.
The following formula was used for the NRI calculation:

@@

\begin{figure} [H]
\begin{center}
\makebox[\textwidth]{\includegraphics[width=\textwidth] {spectral-signatures-1.pdf}}
\caption{Spectral signatures (mean and standard deviation) of Laukiz1, Laukiz2, Luiando and Oiartzun.}
\label{fig:spectral-signatures}
\end{center}
\end{figure}

@@

\section{Correlation among filter methods}

\begin{figure} [H]
\begin{center}
\makebox[\textwidth]{\includegraphics[width=\textwidth] {filter-correlation-nri-1.pdf}}
\caption{Spearman correlation of filter rankings between various filter methods. Results of the NRI feature set are shown.}
\label{fig:correlation-filters}
\end{center}
\end{figure}

\section{Effect of different \texttt{$n_{bins}$} values on filter "information gain"}

\begin{figure} [H]
\begin{center}
\makebox[\textwidth]{\includegraphics[width=\textwidth] {correlation-nbins-1.pdf}}
\caption{Spearman correlation of filter information gain using different \texttt{$n_{bins}$} values for discretization of the numeric response.}
\label{fig:correlation-nbins}
\end{center}
\end{figure}
